---
title: "One-to-N sequence matches in coronavirus genomes"
author: "Jacques van Helden"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
    code_folding: hide
    fig_caption: yes
    highlight: zenburn
    self_contained: no
    theme: cerulean
    toc: yes
    toc_depth: 3
    toc_float: yes
  ioslides_presentation:
    colortheme: dolphin
    fig_caption: yes
    fig_height: 6
    fig_width: 7
    fonttheme: structurebold
    highlight: tango
    smaller: yes
    toc: yes
    widescreen: yes
  beamer_presentation:
    colortheme: dolphin
    fig_caption: yes
    fig_height: 6
    fig_width: 7
    fonttheme: structurebold
    highlight: tango
    incremental: no
    keep_tex: no
    slide_level: 2
    theme: Montpellier
    toc: yes
  word_document:
    toc: yes
    toc_depth: '3'
  slidy_presentation:
    fig_caption: yes
    fig_width: 7
    highlight: tango
    incremental: no
    keep_md: yes
    smaller: yes
    theme: cerulean
    toc: yes
    widescreen: yes
  pdf_document:
    fig_caption: yes
    highlight: zenburn
    toc: yes
    toc_depth: 3
font-import: http://fonts.googleapis.com/css?family=Risque
font-family: Garamond
transition: linear
---

```{r libraries, echo=FALSE, results=FALSE, warning=FALSE, message=FALSE}
#### Install required packages ####
required.packages <- c("knitr")

for (pkg in required.packages) {
  if (!require(pkg, character.only = TRUE)) {
    message("Installing package ", pkg)
    install.packages(pkg, dependencies = TRUE)
  }
  require(pkg, character.only = TRUE)
}

#### Load libraries ####
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}
if (!require("Biostrings", quietly = TRUE)) {
  BiocManager::install("Biostrings")
  require("Biostrings")
}


```



```{r knitr_settings, include=FALSE, echo=FALSE, eval=TRUE}
library(knitr)
options(width = 300)
knitr::opts_chunk$set(
  fig.width = 7, fig.height = 5, 
  fig.path = 'figures/one-to-n-matches/',
  fig.align = "center", 
  size = "tiny", 
  echo = TRUE, eval = TRUE, 
  warning = FALSE, message = FALSE, 
  results = TRUE, comment = "")
# knitr::asis_output("\\footnotesize")


## Store original graphic parameters to restore them after chunks
par.ori <- par(no.readonly = TRUE)



```


```{r parameters}

#### General parameters for the analysis ####

reloadImage <- FALSE

#### Sequence collection ####

## Supported: 
collections <- c(
  "around-CoV-2", 
  "selected",
  "around-CoV-2-plus-GISAID", 
  "selected-plus-GISAID"
  )



# collection <- "around-CoV-2" # 14 genomes
# collection <- "around-CoV-2-plus-GISAID" # 16 genomes
collection <- "all-plus-GISAID" # 16 genomes
# collection <- "selected" # ~60 genomes
# collection <- "selected-plus-GISAID" # ~60 genomes

## Note about GIDAID sequences.
## 
## A few genomes were not available in NCBI Genbank at the time of 
## this analysis, and had to be downloaded from GISAID. These sequences
##  can however not be redistributed, they should thus be downloaded 
##  manually to reproduce the full trees. 


## Exclude incomplete genomes (i.e. those containing a lof of Ns) to avoid biases in the distance computations
excludeIncomplete <- TRUE



#### Define directories and files ####
dir <- list(main = '..')
dir$R <- file.path(dir$main, "scripts/R")

#### Create output directory for sequences ####
dir$seqdata <- file.path(dir$main, "data")
dir.create(dir$seqdata, showWarnings = FALSE, recursive = TRUE)

## Instantiate a list for output files
outfiles <- vector()

## Memory image
dir$images <- file.path(dir$main, "memory_images")
dir.create(dir$images, recursive = TRUE, showWarnings = FALSE)
outfiles["Memory image"] <- file.path(
  dir$images, 
  paste0(collection, "_one-to-n_matches.Rdata"))

## Input files
infiles <- list()

## Output tables
# di$output <- file.path(dir.main, "")
# dir$tables <- 

## Load custom functions
source(file.path(dir$R, "align_n_to_one.R"))
source(file.path(dir$R, "plot_pip_profiles.R"))

## A unequivocal pattern to identify the reference genome in the sequence names of the input file
refPattern <- "HuCoV2_WH01_2019"

## Exclude some genomes with a lot of Ns, because they bias the PIP profiles and alignments and trees
excludePatterns <- c("PnMP789", "PnGu-P2S_2019")

#### Features of interest in the reference genome ####

features <- list()


## Specific features 


## S1: pre-cleavage part of the spike protein 
features[["S1"]] <- c(start = 21599, end =	23617)
## S2: post-cleavage part of the spike protein 
features[["S2"]] <- c(start = 23618, end = 25381)
## Receptor Binding Domain (RBD)
features[["RBD"]] <- c(start = 22517, end = 23185)
## Potential Pangolin origin after Xiao (https://doi.org/10.1101/2020.02.17.951335)
features[['Recomb-Xiao']] <- c(start = 22871, end = 23092)
## Recombinant region 1 seen on the PIP profiles
features[['Recomb-reg-1']] <- c(start = 21500, end = 22800)
## Recombinant region 2 seen on the PIP profiles
features[['Recomb-reg-2']] <- c(start = 22800, end = 24000)
## Recombinant region 3 seen on the PIP profiles
features[['Recomb-reg-3']] <- c(start = 27800, end = 28350)

## Annotated coding sequences
features[['CDS-S']] <- c(start = 21563, end = 25384) ## Spike gene
features[['CDS-ORF3a']] <- c(start = 25393, end = 26220)
features[['CDS-E']] <- c(start = 26245, end = 26472)
features[['CDS-M']] <- c(start = 26523, end = 27191)
features[['CDS-ORF6']] <- c(start = 27202, end = 27387)
features[['CDS-ORF7a']] <- c(start = 27394, end = 27759)
features[['CDS-ORF8']] <- c(start = 27894, end = 28259)
features[['CDS-N']] <- c(start = 28274, end = 29533)
features[['CDS-ORF10']] <- c(start = 29558, end = 29674)
features[['CDS-ORF1ab']] <- c(start = 266, end = 21555)

## All the sequences after the bif ORF coding for the polyproteinn 1ab
features[['After-ORF1ab']] <- c(start = 21556, end = 29899)



## Report the parameters
message("\tReference genomes: ", refPattern)

```


```{r load_sequences}


## Genome dir and files
if (length(grep(pattern = "GISAID", x = collection)) > 0) {
  useGISAID <- TRUE
  dir$genomes <- file.path(dir$main, "data", "GISAID_genomes")
  # collections <-  paste0(collections, "-plus-GISAID")
  # collection <-  paste0(collection, "-plus-GISAID")
} else {
  dir$genomes <- file.path(dir$main, "data", "genomes")
}


## Define the input genome
infiles$genomes <- file.path(
  dir$genomes, 
  paste0("genomes_", collection, ".fasta"))

## Genome sequences
if (!file.exists(infiles$genomes)) {
  stop("Genome sequence file is missing", "\n", infiles$genomes)
}

#### Load genome sequences ####
genomes <- readDNAStringSet(filepath = infiles$genome, format = "fasta")

## Shorten sequence names by suppressing the fasta comment (after the space)
names(genomes) <- sub(pattern = " .*", replacement = "", x = names(genomes), perl = TRUE)

## Exclude genomes
if (excludeIncomplete) {
  excludePattern = paste0("(", paste(collapse = ")|(", excludePatterns), ")")
  excludedGenomeNames <- grep(pattern = excludePattern, x = names(genomes), 
                              value = TRUE, invert = FALSE)
  filteredGenomeIndices <- grep(pattern = excludePattern, x = names(genomes), 
                                value = FALSE, invert = TRUE)
  message("\tExcluded ", length(excludedGenomeNames)," genomes: ", paste(collapse = ", ", excludedGenomeNames)) 
  message("\tRemaining genomes: ", length(filteredGenomeIndices))
  genomes <- genomes[filteredGenomeIndices]
# names(genomes)
}

## Report the number of genoomes
genomeNames <- names(genomes)
nbGenomes <- length(genomeNames)
message("\tLoaded ", nbGenomes, " genomes from file ", infiles$genomes)
# View(genomes)


#### Define reference and query genomes ####
refGenomeName <- grep(pattern = refPattern, x = names(genomes), 
                      ignore.case = TRUE, value = TRUE)
if (is.null(refGenomeName)) {
  stop("Could not identify reference genome with pattern ", refPattern)
}
message("\tReference genome name: ", refGenomeName)

## Compute some statistics about genome sizes
genomeStat <- data.frame(
  n = 1:length(genomes),
  row.names = names(genomes),
  status = rep("Query", length.out = length(genomeNames))
)
genomeStat[,"status"] <- as.vector(genomeStat[,"status"])
genomeStat[refGenomeName,"status"] <- "Reference"
g <- 1
for (g in genomeNames) {
  genomeStat[g, "length"] <- length(genomes[[g]])
}

```



```{r sequence_colors}
#### Define the color associated to each sequence ####

## Color palette per species
speciesPalette <- list(
  Human = "#880000",
  Bat = "#888888",
  Pangolin = "#448800",
  Camel = "#BB8800",
  Pig = "#FFBBBB",
  Civet = "#00BBFF"
)

## Species prefix in the tip labels
speciesPrefix <- c("Hu" = "Human",
                   "Bt" = "Bat",
                   "Pn" = "Pangolin",
                   "Cm" = "Camel",
                   "Pi" = "Pig",
                   "Cv" = "Civet")

## Strain-specific colors
strainColor <- c(
  "HuCoV2_WH01_2019" = "red",
  "HuSARS-Frankfurt-1_2003" = "#0044BB",
  "PnGu1_2019" = "#00BB00",
  "BtRaTG13_" = "#FF6600",
  "BtYu-RmYN" = "#FFBB22",
  "BtZXC21" = "black", 
  "BtZC45" = "black")

## Identify species per tip
for (prefix in names(speciesPrefix)) {
  genomeStat[grep(pattern = paste0("^", prefix), x = row.names(genomeStat), perl = TRUE), "species"] <- speciesPrefix[prefix]
  
}

## Assign acolor to each species
genomeStat$color <- "grey" # default
genomeStat$color <- speciesPalette[as.vector(genomeStat$species)]

for (strain in names(strainColor)) {
  genomeStat[grep(pattern = paste0("^", strain), 
                  x = row.names(genomeStat), perl = TRUE), "color"] <- strainColor[strain]
  
}


## Assign specific color to some nodes

## Define a color for each genome
genomeColors <- (unlist(genomeStat$color))
names(genomeColors) <- row.names(genomeStat)


```

## Parameters

```{r print_parameters}

## Define a list of parameters
parameters <- list()
parameters$collection <- collection
parameters$refGenomeName <- refGenomeName
parameters$nbGenomes <- nbGenomes
parameters$genomeDir <- dir$genomes
parameters$genomeFile <- infiles$genomes
kable(t(as.data.frame.list(parameters)), caption = "Parameters of the analysis", 
      col.names = "Parameter")


```


## Genome statistics

```{r print_genome_stats}
kable(genomeStat, caption = "Reference and query genomes")

```


The collection `r collection` contains `r length(genomes)` virus genome sequences.



## One-to-N alignemnts of selected features

We perform a global pairwise alignment (Needle-Waterman algorithm) between each feature of the reference (`r refGenomeName`) and each one of the query genomes.

```{r one2N_feature_matches, fig.width=10, fig.height=6, out.width="100%", fig.cap="Feature-specific Percent Identical Positions (PIP) profiles. "}
#### One-to-N alignmemnt of user-speficied genomic features ####
# featureName <- "CDS-S" # for the test
featureName <- "S1" # for the test
# featureName <- "Recomb-reg-3" # for the test
# featureName <- "RBD" # for the test
# for (collection in (collections)) {
if (reloadImage) {
  load(outfiles["Memory image"])
  reloadImage <- TRUE

} else {
  allFeatureAlignments <- list()
}
for (featureName in names(features)) {
  featureLimits <- features[[featureName]]
  featureStart <- featureLimits[["start"]]
  featureEnd <- featureLimits[["end"]]
  featureLength <- 
    featureEnd - featureStart  + 1
  
  message("Searching matches for feature ", featureName, 
          " (", featureLimits[1], "-", featureLimits[2], ")", 
          " in collection ", collection)
  
  #### N-to-1 alignments of spike-coding sequences ####
  featurePrefix <- paste0(
    featureName,
    "_", collection)
  
  dir[[featureName]] <- file.path(dir$seqdata, featureName)
  dir.create(dir[[featureName]], showWarnings = FALSE, recursive = TRUE)
  outfiles[featureName]  <- file.path(
    dir[[featureName]], paste0(featurePrefix, ".fasta"))
  message("\tOutput directory: ", dir[[featureName]] )
  
  ## Get sequence of the feature from the reference genome
  refSeq <- subseq(genomes[refGenomeName], 
                   start = featureLimits[1], 
                   end = featureLimits[2])
  
  ## Match the reference feature with all the genomes
  if (reloadImage) {
    featureAlignmentsNto1 <- allFeatureAlignments[[featureName]]
  } else {  
    featureAlignmentsNto1 <- alignNtoOne(
      refSequence = refSeq, 
      querySequences = genomes, 
      type = "global-local",
      outfile = outfiles[featureName])
    
    allFeatureAlignments[[featureName]] <- featureAlignmentsNto1
  }
  
  genomeOrder <- order(featureAlignmentsNto1$stat$score, decreasing = TRUE)
  
  if (featureLength > 20000) {
    windowSize <- 800 
  } else if (featureLength > 5000) {
    windowSize <- 500 
  } else if (featureLength > 3000) {
    windowSize <- 200 
  } else {
    windowSize <- max(100, 10 * round(featureLength / 100))
  }
  
  ## PIP profile of one-to-N alignment
  plotPIPprofiles(
    alignments = featureAlignmentsNto1$alignments[genomeOrder],
    reversePlot = TRUE,
    windowSize = windowSize, 
    main = paste0(featureName, " (", featureStart, ":", featureEnd ,", ", featureLength, " bp)"), 
    #    colors = NULL,
    colors = genomeColors,
    legendMargin = 0.25, 
    legendCorner = "topright", lwd = 1,
    legendCex = 0.5, ylim = c(30, 100))
  
  
  kable(featureAlignmentsNto1$stats[genomeOrder,], 
        caption = paste0(
          "One-to-N alignment of feature ", 
          featureName))
  
}
# }
```


<!-- ```{r one2N_feature_matches_PIP, fig.width=10, fig.height=6, out.width="100%", fig.cap="Feature-specific Percent Identical Positions (PIP) profiles. "} -->
<!-- #### One-to-N alignmemnt of user-speficied genomic features #### -->
<!-- allFeatureAlignments <- list() -->
<!-- # featureName <- "CDS-S" # for the test -->
<!-- featureName <- "S1" # for the test -->
<!-- # featureName <- "Recomb-reg-3" # for the test -->
<!-- # featureName <- "RBD" # for the test -->
<!-- # for (collection in (collections)) { -->
<!-- for (featureName in names(features)) { -->
<!--   featureLimits <- features[[featureName]] -->
<!--   featureStart <- featureLimits[["start"]] -->
<!--   featureEnd <- featureLimits[["end"]] -->
<!--   featureLength <-  -->
<!--      featureEnd - featureStart  + 1 -->

<!--   message("Searching matches for feature ", featureName,  -->
<!--           " (", featureLimits[1], "-", featureLimits[2], ")",  -->
<!--           " in collection ", collection) -->

<!--   #### N-to-1 alignments of spike-coding sequences #### -->
<!--   featurePrefix <- paste0( -->
<!--     featureName, -->
<!--     "_", collection) -->

<!--   dir[[featureName]] <- file.path(dir$seqdata, featureName) -->
<!--   dir.create(dir[[featureName]], showWarnings = FALSE, recursive = TRUE) -->
<!--   outfiles[featureName]  <- file.path( -->
<!--     dir[[featureName]], paste0(featurePrefix, ".fasta")) -->
<!--   message("\tOutput directory: ", dir[[featureName]] ) -->

<!--   ## Get sequence of the feature from the reference genome -->
<!--   refSeq <- subseq(genomes[refGenomeName],  -->
<!--                    start = featureLimits[1],  -->
<!--                    end = featureLimits[2]) -->

<!--   featureAlignmentsNto1 <- allFeatureAlignments[[featureName]] -->

<!--   genomeOrder <- order(featureAlignmentsNto1$stat$score, decreasing = TRUE) -->

<!--   if (featureLength > 20000) { -->
<!--     windowSize <- 800  -->
<!--   } else if (featureLength > 5000) { -->
<!--     windowSize <- 500  -->
<!--   } else if (featureLength > 3000) { -->
<!--     windowSize <- 200  -->
<!--   } else { -->
<!--     windowSize <- max(100, 10 * round(featureLength / 100)) -->
<!--   } -->

<!--   ## PIP profile of one-to-N alignment -->
<!--   plotPIPprofiles( -->
<!--     alignments = featureAlignmentsNto1$alignments[genomeOrder], -->
<!--     reversePlot = TRUE, -->
<!--     windowSize = windowSize,  -->
<!--     main = paste0(featureName, " (", featureStart, ":", featureEnd ,", ", featureLength, " bp)"),  -->
<!-- #    colors = NULL, -->
<!--     colors = genomeColors, -->
<!--     legendMargin = 0.25,  -->
<!--     legendCorner = "topright", lwd = 1, -->
<!--     legendCex = 0.5, ylim = c(30, 100)) -->


<!--   kable(featureAlignmentsNto1$stats[genomeOrder,],  -->
<!--         caption = paste0( -->
<!--           "One-to-N alignment of feature ",  -->
<!--           featureName)) -->

<!-- } -->
<!-- # } -->
<!-- ``` -->

## Output files

```{r output_files}
kable(t(as.data.frame(dir)), col.names = "Dir", caption = "Directories")

outfileTable <- data.frame(path = as.vector(outfiles))
outfileTable$basename <- basename(as.vector(outfileTable$path))
outfileTable$dir <- dirname(as.vector(outfileTable$path))
outfileTable$link <- paste0(
  "[", outfileTable$basename, "](", outfileTable$path, ")"
)

kable(outfileTable[, c("dir", "link")], 
      col.names = c("dir", "file"), 
      caption = "Output files")

```

## Memory image

We store the result ina memory image, in oder to be able reloading it to plot PIP profiles with different parameters. 

```{r}
save.image(file = outfiles["Memory image"])

```


## Session info

```{r session_info}
sessionInfo()
```



