---
title: "PIP profiles of coronavirus spike proteins - DNA and protein sequences"
author: "Jacques van Helden"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
    code_folding: hide
    fig_caption: yes
    highlight: zenburn
    self_contained: no
    theme: cerulean
    toc: yes
    toc_depth: 3
    toc_float: yes
  ioslides_presentation:
    colortheme: dolphin
    fig_caption: yes
    fig_height: 6
    fig_width: 7
    fonttheme: structurebold
    highlight: tango
    smaller: yes
    toc: yes
    widescreen: yes
  beamer_presentation:
    colortheme: dolphin
    fig_caption: yes
    fig_height: 6
    fig_width: 7
    fonttheme: structurebold
    highlight: tango
    incremental: no
    keep_tex: no
    slide_level: 2
    theme: Montpellier
    toc: yes
  word_document:
    toc: yes
    toc_depth: '3'
  slidy_presentation:
    fig_caption: yes
    fig_height: 6
    fig_width: 7
    highlight: tango
    incremental: no
    keep_md: yes
    smaller: yes
    theme: cerulean
    toc: yes
    widescreen: yes
  pdf_document:
    fig_caption: yes
    highlight: zenburn
    toc: yes
    toc_depth: 3
font-import: http://fonts.googleapis.com/css?family=Risque
font-family: Garamond
transition: linear
---

```{r libraries, echo=FALSE, results=FALSE, warning=FALSE, message=FALSE}
#### Install required packages ####
required.packages <- c("knitr")

for (pkg in required.packages) {
  if (!require(pkg, character.only = TRUE)) {
    message("Installing package ", pkg)
    install.packages(pkg, dependencies = TRUE)
  }
  require(pkg, character.only = TRUE)
}

#### Load libraries ####
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}
if (!require("Biostrings", quietly = TRUE)) {
  BiocManager::install("Biostrings")
  require("Biostrings")
}


```



```{r knitr_settings, include=FALSE, echo=FALSE, eval=TRUE}
library(knitr)
options(width = 300)
knitr::opts_chunk$set(
  fig.width = 7, fig.height = 5, 
  fig.path = 'figures/pip_profiles/prot_',
  fig.align = "center", 
  size = "tiny", 
  echo = TRUE, eval = TRUE, 
  warning = FALSE, message = FALSE, 
  results = TRUE, comment = "")
# knitr::asis_output("\\footnotesize")


## Store original graphic parameters to restore them after chunks
par.ori <- par(no.readonly = TRUE)



```


```{r parameters}

#### General parameters for the analysis ####

## Use (or not) GIDAID sequences
## 
## A few genoes were not available in NCBI Genbank at the time of 
## this analysis, and had to be downloaded from GISAID. These sequences
##  can however not be redistributed, they should thus be downloaded 
##  manually to reproduce the full trees. Alternatively, useGISAID
##  can be set to FALSE, whcih will reproduce the analysis with almost 
##  all the sequences of the paper.
useGISAID <- TRUE


#### Define directories and files ####
dir <- list(main = '..')
dir$R <- file.path(dir$main, "scripts/R")

#### Create output directory for sequences ####
seqPrefix <- "spike_proteins"
dir$outseq <- file.path(
  dir$main, "results", seqPrefix, "PIP_profiles")
dir.create(dir$outseq, showWarnings = FALSE, recursive = TRUE)
list.files(dir$outseq)

## Instantiate a list for output files
outfiles <- vector()

## Input files
infiles <- list()

## Sequence collection
## Supported: 
## collection <- "around-CoV-2" # ~20 sequences
## collection <- "selected" # ~60 sequences
## collection <- "all" # ~60 sequences
collections <- c("around-CoV-2", "selected", "all")
# collection <- "around-CoV-2" # ~20 sequences
# collection <- "selected" # ~40 sequences
collection <- "all" # ~60 sequences


## Sequence dir and files
if (useGISAID) {
  dir$sequences <- file.path(dir$main, "data", "GISAID_genomes")
  collection <- paste0(collection, "-plus-GISAID")
} else {
  dir$sequences <- file.path(dir$main, "data", "spike_proteins")
}

seqPrefix <- paste0("spike_proteins_", collection)

infiles$sequences <- file.path(dir$sequences, paste0(seqPrefix,".fasta"))

## Sequence sequences
if (!file.exists(infiles$sequences)) {
  stop("¨Protein sequence file is missing", "\n", infiles$sequences)
}


## Load custom functions
source(file.path(dir$R, "align_n_to_one.R"))
source(file.path(dir$R, "plot_pip_profiles.R"))


## Query patterns for SARS-CoV-2
queryPatterns <- list()
queryPatterns[["HuCoV2_WH01_2019"]] <- c(
  "PnGX-P1E_2017", 
  "PnGu1_2019",
  "BtRaTG13_2013_Yunnan", 
  "BtZC45",
  "BtZXC21",
  "HuSARS-Frankfurt-1_2003",
  "CmMERS",
  "HuMERS_172-06_2015"
)

queryPatterns[["HuSARS-Frankfurt-1_2003"]] <- c(
  "CvSZ3",
  "BtRs4874",
  "BtWIV16_2013",
  "BtRs4231",
  "BtRs7327",
  "BtRsSHC014",
  "Btrec-SARSp_2008",
  "PnGX-P1E_2017",
  "HuCoV2_WH01_2019",
  "BtRaTG13_2013_Yunnan", 
  "BtZC45",
  "BtZXC21",
  "CmMERS",
  "HuMERS_172-06_2015"
)

#### Add GISAID IDs to the query pattern ####
## Note that GISAID sequences are be submitted to the github repo because they cannot be redistributed
if (useGISAID) {
  for (ref in names(queryPatterns))
  queryPatterns[[ref]] <- append(queryPatterns[[ref]], 
                          c("BtYu-RmYN02_2019",
                            "PnGu1_2019"
                          ))
}

# message("\tReference strain: ", refPattern)
message("\tQuery patterns")
for (ref in names(queryPatterns)) {
  queryPatterns[[ref]] <- unique(queryPatterns[[ref]])
  message("\t", ref, "\t\t", length(queryPatterns[[ref]]),
          "\t", paste(collapse = ", ", queryPatterns[[ref]]))
}

```



```{r load_sequences}

#### Load sequences ####
sequences <- readAAStringSet(filepath = infiles$sequences, format = "fasta")

## Shorten sequence names by suppressing the fasta comment (after the space)
names(sequences) <- sub(pattern = " .*", replacement = "", x = names(sequences), perl = TRUE)

sequencesNames <- names(sequences)
nbsequences <- length(sequencesNames)
message("Loaded ", nbsequences, " sequences from file ", infiles$sequences)
# View(sequences)

#### Define reference and query sequences ####
refSequenceNames <- vector()
for (ref in names(queryPatterns)) {
  refSequenceNames[ref] <- unique(
    grep(pattern = ref, x = names(sequences), 
         ignore.case = TRUE, value = TRUE))
    if (is.null(refSequenceNames[ref])) {
    stop("Could not identify reference sequences with pattern ", ref)
  }
}

## Query sequences
querySequenceNames <- list()
for (ref in names(queryPatterns)) {
#  message("Identifying query sequences for reference: ", ref)
  queryRegExp <- paste0("(", paste(collapse = ")|(", queryPatterns[[ref]]), ")")
  querySequenceNames[[ref]] <- grep(pattern = queryRegExp, 
                             x = sequencesNames, 
                             ignore.case = TRUE, value = TRUE)
  nbquerySeq <- length(querySequenceNames[[ref]])
  
  if (nbquerySeq == 0) {
    stop("Could not identify any query sequences with query pattern\n", queryRegExp)
  }
  
  if (length(unlist(queryPatterns[ref[]])) != length(querySequenceNames[[ref]])) {
    foundPatterns <- grep(pattern = queryRegExp, x = querySequenceNames[[ref]], value = TRUE)
    missingPatterns <- setdiff(foundPatterns, queryPatterns[[ref]])
    message("\t", 
            length(missingPatterns), " Missing sequences: ", 
            paste(collapse = ", ", missingPatterns))
  }
  
  message("\tFound ", 
          "\t", length(querySequenceNames[[ref]]), 
          " queries for\t", ref)

  ## Compute some statistics about sequences sizes
  sequencestat <- data.frame(
    row.names = c(ref, querySequenceNames[[ref]]),
    status = c("Reference", 
               rep("Query", length.out = length(querySequenceNames[[ref]])))
  )
  
  g <- 1
  for (g in c(ref, querySequenceNames[[ref]])) {
    sequencestat[g, "length"] <- length(sequences[[g]])
  }
  kable(sequencestat, caption = paste0("Query sequences for reference ", ref))
}


```

The collection `r collection` contains `r length(sequences)` virus sequences sequences.



## N-to-1 full sequence alignments

We perform a pairwise lignment between each sequences query and the reference sequences (`r refSequenceNames[[1]]`).

```{r full-sequences_align}
sequencesNto1 <- list()
refSequenceName <- "HuCoV2_WH01_2019"
for (refSequenceName in refSequenceNames) {
  ## Define output file for sequences alignments
  outfile <- file.path(
    dir$outseq, paste0("one-to-n_alignments_ref_", refSequenceName))
  outfiles[paste0(refSequenceName, " alignments")] <- outfile
  
  ## Get sequences for reference and query sequences
  refSequence <- sequences[refSequenceName]
  querySeq <- sequences[querySequenceNames[[refSequenceName]]]
  
  message("\tAligning ", length(querySeq), " query sequences",
          " to reference\t", refSequenceName)

  sequencesNto1[[refSequenceName]] <- alignNtoOne(
    refSequence = refSequence,  
    querySequences = querySeq, 
    sortByPIP = TRUE,
    #  querySequences = sequences, 
    outfile = outfile)
  
}

```


## Alignment with SARS-CoV-2(ref strain HuCoV2_WH01_2019)


```{r PIP_SARS-CoV-2, fig.width=12, fig.height=8, out.width="100%", fig.cap="Percent Identical Positions profile over the whole sequences of SARS-CoV-2. "}
refSequenceName <- "HuCoV2_WH01_2019"
kable(sequencesNto1[[refSequenceName]]$stats[order(sequencesNto1[[refSequenceName]]$stats$score, decreasing = TRUE), ], 
        caption = "N-to-one alignment of full sequences")


## PIP profile of full sequences N-to-1 alignments
plotPIPprofiles(
  alignments = sequencesNto1[[refSequenceName]]$alignments, 
  windowSize = 100, 
  main = paste0("Full sequences PIP profile", "\nRef: ", refSequenceName),
  legendMargin = 0, 
  legendCorner = "bottomright", legendCex = 0.6, ylim = c(0,100))

```



## Alignment with SARS-CoV (ref strain: HuSARS-Frankfurt-1_2003)



```{r PIP_SARS-CoV, fig.width=12, fig.height=8, out.width="100%", fig.cap="Percent Identical Positions profile over the whole sequences of SARS-CoV. "}
refSequenceName <- "HuSARS-Frankfurt-1_2003"
kable(sequencesNto1[[refSequenceName]]$stats[order(sequencesNto1[[refSequenceName]]$stats$score, decreasing = TRUE), ], 
        caption = "N-to-one alignment of full sequences")


## PIP profile of full sequences N-to-1 alignments
plotPIPprofiles(alignments = sequencesNto1[[refSequenceName]]$alignments, 
                windowSize = 100, 
                #                legend = paste0(names(sequencesNto1[[refSequenceName]]$alignments), " (", round(digits = 2, sequencesNto1[[refSequenceName]]$stats$pid), "%)"),
                main = paste0("Full sequences PIP profile", "\nRef: ", refSequenceName),
                legendMargin = 0, 
                legendCorner = "bottomright", legendCex = 0.6, ylim = c(0,100))

```


## Aligning all sequences

```{r SARS_vs_all}

## Define output file for sequences alignments
outfiles["Sequence alignments - SARS"] <- file.path(
  dir$outseq, paste0("spike-protein_alignments_ref_",
                     "HuSARS-Frankfurt-1_2003"))

#### Compare SARS (2002) wih the related Civet sequences ####
SARSvsAll <- alignNtoOne(
  refSequence = sequences["HuSARS-Frankfurt-1_2003"], 
  querySeq = sequences, sortByPIP = TRUE,
  outfile = outfiles["Sequence alignments - SARS"] )

kable(SARSvsAll$stats, 
      caption = "SARS (2003). N-to-one alignment of full sequences of the closest animal virus (Civet) and of Human SARS-CoV-2. ")

```

```{r PIP_SARS_vs_all, fig.width=10, fig.height=5, out.width="100%", fig.cap="Percent Identical Positions profile over the whole sequences of SARS (2002-2003). "}

## PIP profile of full sequences N-to-1 alignments
plotPIPprofiles(alignments = SARSvsAll$alignments, 
                windowSize = 500, 
                legend = paste0(names(SARSvsAll$alignments), " (", round(digits = 2, SARSvsAll$stats$pid), "%)"),
                main = paste0("Spike protein PIP", "\nRef: ", "Human_SARS-CoV_Frankfurt_1"),
                legendMargin = 0.3, 
                legendCorner = "bottomright", 
                legendCex = 0.5, 
                ylim = c(0,100))


```




<!-- ## Spike protein against 2003 SARS-CoV (Frankfurt strain) -->

<!-- ```{r S-gene_align_queries} -->

<!-- featureName <- "Spike-protein_SARS-CoV" -->
<!-- featureLimits <- features[[featureName]] -->


<!-- #### N-to-1 alignments of spike-coding sequences #### -->
<!-- dir[[featureName]] <- file.path(dir$main, "results", featureName) -->
<!-- dir.create(featureName, showWarnings = FALSE, recursive = TRUE) -->
<!-- outfiles[featureName]  <- file.path( -->
<!--   dir[[featureName]], paste0(featureName, "_", collection, "_alignments")) -->

<!-- ## Get sequences for reference and query sequences -->
<!-- refSeq <- subseq(sequences[refSequenceName], start = featureLimits[1], end = featureLimits[2]) -->

<!-- featureAlignmentsNto1 <- alignNtoOne( -->
<!--   refSequence = refSeq,  -->
<!--   querySeq = querySeq,  -->
<!--   type = "global-local", sortByPIP = TRUE, -->
<!--   outfile = outfiles[featureName]) -->

<!-- kable(featureAlignmentsNto1$stats[order(featureAlignmentsNto1$stat$score, decreasing = TRUE),],  -->
<!--       caption = "N-to-one alignment of S genes") -->

<!-- ``` -->


<!-- ### Spike gene PIP plot -->

<!-- ```{r Sgene_pip, fig.width=10, fig.height=5, out.width="100%", fig.cap="Percent Identical Positions profile over the whole sequences of SARS-CoV-2. "} -->

<!-- ## PIP profile of spike N-to-1 alignments -->
<!-- plotPIPprofiles(alignments = featureAlignmentsNto1$alignments,  -->
<!--                 windowSize = 200,  -->
<!-- #                legend = paste0(names(featureAlignmentsNto1$alignments), " (", round(digits = 2, featureAlignmentsNto1$stats$pid), "%)"), -->
<!--                 main = paste0(featureName, " - PIP profile", "\nRef: ", refSequenceName), -->
<!--                 legendMargin = 0, legendCorner = "bottomright", legendCex = 0.8, ylim = c(30, 100)) -->

<!-- ``` -->




<!-- ### Getting sequences of spike gene in all the selected coronavirus sequences -->


<!-- ```{r  match sequence fragment against all sequences and export the matches to a fasta file} -->


<!-- #' @title Export sequencesquery + subject sequences from N-to-1 alignments -->
<!-- #' @author Jacques.van-Helden@france)bioinformatique.fr -->
<!-- #' @param refsequence  sequence used as reference (DB) for the N-to-1 alignment -->
<!-- #' @param sequences sequences sequences -->
<!-- #' @param IDsuffix=NULL if not null, append the suffix to the sequence name for the fasta header -->
<!-- #' @param outfile=NULL output file. If specified, the matches will be exported as fasta-formatted sequences. -->
<!-- #' @export -->
<!-- ExportSegmentMatches <- function(refsequence, -->
<!--                                  sequences, -->
<!--                                  IDsuffix = NULL, -->
<!--                                  outfile=NULL) { -->

<!--   #### N-to-1 alignments of all spike-coding sequences #### -->

<!--   ## Get sequences for reference and query sequences -->
<!--   alignmentsNto1 <- alignNtoOne( -->
<!--     refSequence = refsequence,  -->
<!--     querySeq = sequences,  -->
<!--     type = "global-local") -->

<!--   if (!is.null(outfile)) { -->

<!--     ## Write the reference sequence in the output fle -->
<!--     writeXStringSet(refsequence, filepath = outfile, format = "fasta") -->

<!--     i <- 1 -->
<!--     nbAlignments <- length(alignmentsNto1$alignments) -->
<!--     for (i in 1:nbAlignments) { -->
<!--       sequencesName <- names(sequences[i]) -->
<!--       alignment <- alignmentsNto1$alignments[[i]] -->
<!--       subject <- subject(alignment) -->

<!--       ## Suppress the dashes from the alignment to get the raw sequence -->
<!--       sequence <- as.character(subject) -->
<!--       sequenceDesaligned <- gsub(pattern = "-", replacement = "", x = sequence) -->
<!--       seqStringSet <- DNAStringSet(x = sequenceDesaligned) #, start = start(subject), end=end(subject)) -->


<!--       ## Define a sequence ID for the fasta header -->
<!--       sequenceID <- sequencesName -->
<!--       if (!is.null(IDsuffix)) { -->
<!--         sequenceID <- paste0(sequenceID, IDsuffix) -->
<!--       }  -->
<!--       sequenceID <- paste0(sequenceID, "_", start(subject), "-", end(subject)) -->
<!--       names(seqStringSet) <- sequenceID -->
<!--       #  -->
<!--       ## Write pairwise alignment (temporarily disaactivated) -->
<!--       # alignmentFile <- paste0("pairwise-alignment_",  -->
<!--       #                         # gsub(pattern = "/", replacement = "-", x = sequencesName),  -->
<!--       #                         ".txt") -->
<!--       # writePairwiseAlignments(x = alignment, file = outfile) -->

<!--       ## Append the sequence to the file -->
<!--       message("\tAppending sequence ", i, "/", nbAlignments, "\t", sequenceID) -->
<!--       writeXStringSet(seqStringSet, -->
<!--                       filepath = outfile, format = "fasta", append = TRUE) -->

<!--     } -->
<!--     message("\tExported alignments to\t", outfile) -->

<!--   }  -->
<!--   return(alignmentsNto1) -->

<!-- } -->
<!-- ``` -->

<!-- ```{r Sgene_exportseq} -->


<!-- #### Export S gene sequences found in the reference sequences #### -->
<!-- # toString(refSeq) -->
<!-- writeXStringSet( -->
<!--   refSeq,  -->
<!--   filepath = file.path( -->
<!--     dir$outseq,  -->
<!--     paste0("S-gene_", refSequenceName, ".fasta")),  -->
<!--   format = "fasta") -->


<!-- ## Export matches to S-gene -->
<!-- outfiles["S gene matches"] <- file.path( -->
<!--   dir$outseq,  -->
<!--   paste0("S-gene_", refSequenceName, "_matches.fasta")) -->

<!-- featureAlignmentsNto1All <- ExportSegmentMatches( -->
<!--   refsequence = refSeq, -->
<!--   sequences = sequences,  -->
<!--   IDsuffix = "_S-gene", -->
<!--   outfile = outfiles["S gene matches"]) -->


<!-- kable(featureAlignmentsNto1All$stats[order(featureAlignmentsNto1All$stat$score, decreasing = TRUE),],  -->
<!--         caption = "N-to-one alignment of S genes") -->

<!-- ``` -->

<!-- ## Export matches to the coding sequence of the S1 cleavage produc -->


<!-- ## N-to-1 alignemnts of S1 coding sequences -->

<!-- ```{r S1_align_queries} -->

<!-- refS1Limits <- refSpikeLimits[1] + 3*c(13,	685) -->

<!-- #### N-to-1 alignments of S1-coding sequences #### -->

<!-- ## Get sequences for reference and query sequences -->
<!-- refS1 <- subseq(sequences[refSequenceName], start = refS1Limits[1], end = refS1Limits[2]) -->

<!-- S1Nto1 <- alignNtoOne(refSequence = refS1, querySeq = querySeq, type = "global-local") -->

<!-- kable(S1Nto1$stats[order(S1Nto1$stat$score, decreasing = TRUE),],  -->
<!--       caption = "N-to-one alignment of S1 coding sequence") -->

<!-- ``` -->

<!-- ### S1 gene PIP plot -->

<!-- ```{r S1_pip, fig.width=10, fig.height=5, out.width="100%", fig.cap="Percent Identical Positions profile over the whole sequences of SARS-CoV-2. "} -->

<!-- ## PIP profile of S1 N-to-1 alignments -->
<!-- plotPIPprofiles(alignments = S1Nto1$alignments,  -->
<!--                 windowSize = 200,  -->
<!--                 legend = paste0(names(S1Nto1$alignments), " (", round(digits = 2, S1Nto1$stats$pid), "%)"), -->
<!--                 main = paste0("S1 coding sequence - PIP profile", "\nRef: ", refSequenceName), -->
<!--                 legendMargin = 0, legendCorner = "bottomleft", legendCex = 0.8, ylim = c(30, 100)) -->

<!-- ``` -->


## Output files

```{r output_files}
kable(t(as.data.frame(dir)), col.names = "Dir", caption = "Directories")

kable(outfiles, col.names = "File", caption = "Output files")
```

## Session info

```{r session_info}
sessionInfo()
```



